Renormalization

### 2014-05-21 papers/renormalization/paper.tex
Project report writing Tested on elliot. Took 0:32.16.
more renormalization work Tested on bennet. Took 0:18.13.

\subsubsection{Defining our approximation}
Let's begin by considering the partition function and free energy, as
computed in the canonical ensemble:
\begin{align}
  Z &= \sum_i^{\text{\tiny\begin{minipage}{5em}
      \begin{center}
        \tiny all microstates\end{center}\end{minipage}}}
       e^{-\beta E_i}, & F &= -kT\log Z
\end{align}
We now consider a ``smooth'' configuration that contains only
microstates that are periodic with a given periodicity $2^j\lambda$:
\newcommand\sumstatesperiodic[1]{\sum_i^{\text{\tiny\begin{minipage}{6em}
      \begin{center}
        \tiny microstates periodic $#1$\end{center}\end{minipage}}}}
\begin{align}
  Z_j &= \sumstatesperiodic{2^j\lambda}
       e^{-\beta E_i}, & F_j &= -kT\log Z_j
       \label{eq:Zn}
\end{align}
This defines a family of free energies that contain fluctuations with
a variety of length scales.  For the moment we will assume a simple
cubic periodicity, although it may be advantageous to select face
centered cubic (FCC) instead.  So far, we have defined a family of
free energy functions (each containing longer length-scale
fluctuations) that we can compute numerically using Monte Carlo, but
we have not yet constructed a means to approximate these functions.
Before developing this approximation, let's reorganize this sum over
periodic states.
\newcommand\sumoccupanciesofoctants{
  \sum_{N_1\cdots N_8}^{\text{\tiny\begin{minipage}{5em}
      \begin{center}
      \tiny numbers in octants\end{center}\end{minipage}}}
}
\newcommand\summicrostatesofoctants{
  \sum_{i_1\cdots i_8}^{\text{\tiny\begin{minipage}{4em}
      \begin{center}
      \tiny microstates\\within octants\end{center}\end{minipage}}}
}
\begin{align}
  Z_j(N) &= \sumoccupanciesofoctants g_{N_1\cdots N_8}
       \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
where here $g_{N_1\cdots N_8}$ is a multiplicity factor
corresponding to the number of ways the cell can be divided into
subcells with occupancies of $N_1\cdots N_8$.  This is so far only a
different way of writing Eq.~\ref{eq:Zn}.  At this point, we can
multiply and divide by a few factors, with a goal of replacing the
unknown $E_{i_1\cdots i_8}^{(j)}$ with something that is easier to
approximate.
\begin{multline}
  F_j(n) = -kT\log\Bigg(
    \sumoccupanciesofoctants
       g_{N_1\cdots N_8}
       \\
       \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
  \Bigg)
\end{multline}

@@

The flavor of the approximation used by White and others in GRG is now
clarified.  We begin by assuming that we already have an adequate (and
tractable) approximation for $F_{j-1}$.  Thus the denominator
partition function is readily computable.  We further simplify the
numerator by multiplying and dividing by the same quantity, and
simplify the notation by defining
\begin{align}
  F_j(N_1\cdots N_8) &= \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
and thus rewrite the energy difference as
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sumoccupanciesofoctants
       e^{-\beta \big(F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) +
           \big(F_j(N_1\cdots N_8) - F_{j-1}(N_1) - \cdots - F_{j-1}(N_8)\big) \big)}
  }{
    \sumoccupanciesofoctants
       e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \right)}
  }
  \right)
\end{multline}
At this point we can recognize and identify two energy differences as
\begin{align}
  \bar{F}(N_1\cdots N_8) &= F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \\
  \bar{U}(N_1\cdots N_8) &= F_j(N_1\cdots N_8)
  - \left(F_{j-1}(N_1) + \cdots + F_{j-1}(N_8) \right)
\end{align}
and using these definitions can rewrite our energy difference in a
more compact form:
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sumoccupanciesofoctants
       e^{-\beta \big( \bar{F}(N_1\cdots N_8) + \bar{U}(N_1\cdots N_8)
         \big)}
  }{
    \sumoccupanciesofoctants
       e^{-\beta \bar{F}(N_1\cdots N_8)}
  }
  \right)
\end{multline}
At this point, we can recognize $\bar{F}(N_1\cdots N_8)$ as the free
energy change due to oscillations \emph{if we neglect the actual
  energy of the oscillations themselves, but instead make a ``local''
  approximation}.  Meanwhile $\bar{U}(N_1\cdots N_8)$ is the free
energy change due to the energy of oscillations.

\fixme{I have now ``defined'' the terms used by Forte in her
  notation of GRG.  On the to-do list: }
\begin{enumerate}
\item \fixme{Elaborate on the approximation (ignoring corners) that
  leads from a $\sumoccupanciesofoctants g_{N_1\cdots N_8}$ to an
  integral over density amplitudes.}
\item \fixme{Develop an approach for approximating $\bar{U}$.  Perhaps
  is $F_{j-1}$ is expressed as a functional, we can use a non-local
  DFT to approximate $\bar U_j$?  The standard GRG approach seems to
  be to use a simple mean-field or thermodynamic perturbation theory
  approach to approximate $\bar U$, suggesting that a simple
  (standard) DFT might work just fine for this.}
\end{enumerate}

>>>>

\subsubsection{Defining our approximation}
Let's begin by considering the partition function and free energy, as
computed in the canonical ensemble:
\begin{align}
  Z &= \sum_i^{\text{\tiny\begin{minipage}{5em}
      \begin{center}
        \tiny all microstates\end{center}\end{minipage}}}
       e^{-\beta E_i}, & F &= -kT\log Z
\end{align}
We now consider a ``smooth'' configuration that contains only
microstates that are periodic with a given periodicity $2^j\lambda$:
\newcommand\sumstatesperiodic[1]{\sum_i^{\text{\tiny\begin{minipage}{6em}
      \begin{center}
        \tiny microstates periodic $#1$\end{center}\end{minipage}}}}
\begin{align}
  Z_j &= \sumstatesperiodic{2^j\lambda}
       e^{-\beta E_i}, & F_j &= -kT\log Z_j
       \label{eq:Zn}
\end{align}
This defines a family of free energies that contain fluctuations with
a variety of length scales.  For the moment we will assume a simple
cubic periodicity, although it may be advantageous to select face
centered cubic (FCC) instead.  So far, we have defined a family of
free energy functions (each containing longer length-scale
fluctuations) that we can compute numerically using Monte Carlo, but
we have not yet constructed a means to approximate these functions.
Before developing this approximation, let's reorganize this sum over
periodic states.
\begin{figure}
  \centering
  \includegraphics[width=0.8\columnwidth]{figs/divisions}
  \caption{A cell being subdivided with $N_1\cdots N_8$ particles in
    each subdivision.}
\end{figure}
\newcommand\sumoccupanciesofoctants{
  \sum_{N_1\cdots N_8}^{\text{\tiny\begin{minipage}{5em}
      \begin{center}
      \tiny numbers in octants\end{center}\end{minipage}}}
}
\newcommand\summicrostatesofoctants{
  \sum_{i_1\cdots i_8}^{\text{\tiny\begin{minipage}{4em}
      \begin{center}
      \tiny microstates\\within octants\end{center}\end{minipage}}}
}
\begin{align}
  Z_j(N) &= \sumoccupanciesofoctants g_{N_1\cdots N_8}
       \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
where here $g_{N_1\cdots N_8}$ is a multiplicity factor
corresponding to the number of ways the cell can be divided into
subcells with occupancies of $N_1\cdots N_8$.  This is so far only a
different way of writing Eq.~\ref{eq:Zn}.  At this point, we can
multiply and divide by a few factors, with a goal of replacing the
unknown $E_{i_1\cdots i_8}^{(j)}$ with something that is easier to
approximate.
\begin{multline}
  F_j(n) = -kT\log\Bigg(
    \sumoccupanciesofoctants
       g_{N_1\cdots N_8}
       \\
       \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
  \Bigg)
\end{multline}

@@

The flavor of the approximation used by White and others in GRG is now
clarified.  We begin by assuming that we already have an adequate (and
tractable) approximation for $F_{j-1}$.  Thus the denominator
partition function is readily computable.  We further simplify the
numerator by multiplying and dividing by the same quantity, and
simplify the notation by defining
\begin{align}
  F_j(N_1\cdots N_8) &= g_{N_1\cdots N_8}
       \summicrostatesofoctants
       e^{-\beta E_{i_1\cdots i_8}^{(j)}}
\end{align}
\begin{widetext}
and thus rewrite the energy difference as
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sumoccupanciesofoctants
       e^{-\beta \big(F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) +
           \big(F_j(N_1\cdots N_8) - F_{j-1}(N_1) - \cdots - F_{j-1}(N_8)\big) \big)}
  }{
    \sumoccupanciesofoctants
       e^{-\beta \left( F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \right)}
  }
  \right)
\end{multline}
\end{widetext}
At this point we can recognize and identify two energy differences as
\begin{align}
  \bar{F}(N_1\cdots N_8) &= F_{j-1}(N_1) + \cdots F_{j-1}(N_8)  - F_{j-1}(n) \\
  \bar{U}(N_1\cdots N_8) &= F_j(N_1\cdots N_8)
  - \left(F_{j-1}(N_1) + \cdots + F_{j-1}(N_8) \right)
\end{align}
and using these definitions can rewrite our energy difference in a
more compact form:
\begin{multline}
  F_j(n) - F_{j-1}(n) =\\
  -kT\log\left(\frac{
    \sumoccupanciesofoctants
       e^{-\beta \big( \bar{F}(N_1\cdots N_8) + \bar{U}(N_1\cdots N_8)
         \big)}
  }{
    \sumoccupanciesofoctants
       e^{-\beta \bar{F}(N_1\cdots N_8)}
  }
  \right)
\end{multline}
At this point, we can recognize $\bar{F}(N_1\cdots N_8)$ as the free
energy change due to oscillations \emph{if we neglect the actual
  energy of the oscillations themselves, but instead make a ``local''
  approximation}.  Meanwhile $\bar{U}(N_1\cdots N_8)$ is the free
energy change due to the energy of oscillations.

\begin{figure}
  \centering
  \includegraphics[width=0.8\columnwidth]{figs/interactions}
  \caption{Interactions between neighboring subdivisions.}\label{fig:interactions}
\end{figure}

\subsubsection{Interaction between subsystems}
At this stage, there is just one term that need be approximated
(assuming, as always, that we have an acceptable approximation for
$F_{j-1}$), which is $\bar{U}$.  This is the free energy difference
between two free energies.  The first, $F_j(N_1\cdots N_8)$, is the
fully interacting system that is periodic with periodicity
$2^j\lambda$, and which is constrained to have $N_1\cdots N_8$
particles in each of the subdivisions of the cell.  The second is the
sum of $F_{j-1}$ for each of those subdivisions.  This difference
describes the interaction between the octants of the cell, and is at
the heart of the GRG method.

We desire to simplify the summation over seven independent numbers
$N_1\cdots N_7$ (with the eighth constrained by $\sum_i N_i = N$) to a
summation over a single fluctuation amplitude.  As illustrated in
Fig.~\ref{fig:interactions}, these interactions can be seen as
a sum of interactions between neighboring subdivisions, combined with
a many-body correction due to ``corner'' interactions.

\subsection{To do later}

\begin{enumerate}
\item \fixme{Elaborate on the approximation (ignoring corners) that
  leads from a $\sumoccupanciesofoctants g_{N_1\cdots N_8}$ to an
  integral over density amplitudes.}
\item \fixme{Develop an approach for approximating $\bar{U}$.  Perhaps
  is $F_{j-1}$ is expressed as a functional, we can use a non-local
  DFT to approximate $\bar U_j$?  The standard GRG approach seems to
  be to use a simple mean-field or thermodynamic perturbation theory
  approach to approximate $\bar U$, suggesting that a simple
  (standard) DFT might work just fine for this.}
\end{enumerate}
